The imprint of global magma oceans on exoplanet demographics

### 2023-02-22 src/tex/ms.tex
draft discussion on the detectability of the rgh transition

\section{Methods}\label{sec:methods}
\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.

@@

\subsection{Testability of the runaway greenhouse hypothesis}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical detection of the runaway greenhouse threshold.
For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Should a detection be possible, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.

@@

\subsection{Detectability of the runaway greenhouse threshold}
\todo[inline]{the demographic imprint is detectable. Under what conditions?\\
Mention multiplanet systems.}

%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ... \todo[]{double-check which survey params are needed}
We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
\begin{itemize}
    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be. Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space. If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
\end{itemize}
Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).

\subsubsection{False positive scenarios}
\todo{what other processes could be confused with a magma ocean signal?}
\begin{note}
    Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
    Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}

@@

@@

\begin{large}\textit{Author contributions:}\end{large}

>>>>

\section{Methods}\label{sec:methods}
\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}. Mention we only consider instellation-induced rgh.}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical detection of the runaway greenhouse threshold.
For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.

@@

\subsection{Detectability of the runaway greenhouse transition}
\todo[inline]{the demographic imprint is detectable. Under what conditions?\\
Mention multiplanet systems.}
We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
However, even assuming capable instrumentation and an optimal survey design, geophysical and demographic unknowns may jeopardize this undertaking.
The detectability of the runaway greenhouse threshold is thus a function of both survey design and astrophysical circumstances.

%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ... \todo[]{double-check which survey params are needed}
We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
\begin{itemize}
    \item Occurrence rate of planets forming steam atmospheres: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas. This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{how much?}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
    \item Planetary evolution and duration of the steam atmosphere phase: An inflated steam atmosphere can only be sustained until the planet has lost its water. Depending on host star spectral type and planetary composition, planets spend from a few Myr to hundreds of Myr in runaway greenhouse climates~\citep[e.g.,][]{Luger2015}, and only planets observed during this phase will contribute to the signal.
    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
    \item Sample size: Since we are looking for a population-level trend, the statistical significance increases with a larger sample size.
    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be. Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space. If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
\end{itemize}
Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).

\subsubsection{False positive scenarios}
\todo{what other processes could be confused with a magma ocean signal?\\
If there's no other subsubsection here, make this a (the final?) subsection in ``Discussion''.}
\begin{note}
    Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
    Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}

@@

The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
...

@@

\begin{large}\textit{Author contributions:}\end{large}
%M.S., D.A., and T.L. conceived the project.
%M.S. wrote the manuscript and carried out the hypothesis tests and statistical analyses.
%...
%K.H.-U. provided the star generator based on the Gaia catalog.
%G.B. implemented the planet generator in the Bioverse framework.
%All authors provided comments and suggestions on the manuscript.
